# Our first plot

To review we said that plots are made up of layers, scales, coordinate systems, faceting, and themes. We begin by focusing on the layers and relying on defaults for all the rest. Each layer we said is composed of data, aesthetic mappings, geoms, stat and position. We will focus on **data**, **aesthetic mappings**, and **geoms**, and rely on defaults for stat and position.

::: callout-tip
Always start with the geom and ask yourself: what are the required aesthetic mappings?
:::

Let's make a histogram using the built-in dataset `faithful`. We confirm that it's a data frame and then determine that both variables are continuous (type `num`) and therefore either would work with `geom_histogram()`, which as we said requires continuous aesthetic mappings:

```{r}
class(faithful)
str(faithful)
```

Let's draw a histogram with the `waiting` variable. We now have everything we need:

1. data: `faithful`
2. mapping: `x` ➞ `waiting`
3. geom: `geom_histogram()`

All that's left is to convert this to **ggplot2** code. Typically the data and mapping are indicated in the call to `ggplot()` which initializes the plot, and then we add in the geom.

```{r}
#| message: true
library(ggplot2)
ggplot(data = faithful, mapping = aes(x = waiting)) +
  geom_histogram()
```

And we have our first plot! A few things to note:

* You cannot start a new line with <code class="sourceCode r"><span class="sc">+</span></code>.

* It 


```{r}
#| eval: false
# comment
finches <- Sleuth3::case0201
ggplot(finches, aes(x = Depth, y = after_stat(density))) +
  geom_histogram(bins = 20, color = "#80593D", fill = "#9FC29F", alpha = .5) +
  geom_density(color = "#3D6480", lwd = 1) +
  facet_wrap(~Year) +
  labs(title = "Beak Depth Density of Galapagos Finches",
       x = "Beak Depth (mm)", caption = "Source: Sleuth3::case0201") +
  theme_grey(13)
```

